Data Migration under a Changing Schema in Ampersand

### 2023-11-05 2022Migration/articleMigrationFACS.tex
minor

\title{Data Migration under a Changing Schema in Ampersand}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Sebastiaan Joosten\inst{1}\orcidID{0000-0002-6590-6220}\\ \and
Stef Joosten\inst{2,3}\orcidID{0000-0001-8308-0189}}
%
\authorrunning{S. Joosten}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{University of Minnesota, Minneapolis, USA \and Open Universiteit, Heerlen, the Netherlands\\ 
\email{stef.joosten@ou.nl} \and
Ordina NV, Nieuwegein, the Netherlands}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
   Software generators can help increase the frequency of releases and their reliability.
   They save on time spent on development and fixing human-induced mistakes by compiling a specification into a working information system.
   However, many generators do not support data migrations.
   Data migration is necessary when an incremental deployment changes the system's schema.
   Consequently, developers tend to avoid migrations or migrate data ``by hand''.

@@

   This paper defines a migration process that preserves the semantics of data by means of invariants (constraints)
   with zero downtime for users.
   The migration process is based on assumptions and requirements that aim to capture a large class of data migrations in practice.
   This paper focuses on correctness of the migration. Efficiency of the migration is outside its scope.
\keywords{Generative software \and Incremental software deployment \and Data migration \and Relation algebra \and Ampersand \and Schema change. \and Invariants \and Zero downtime.}
\end{abstract}

\section{Introduction}
\label{sct:Introduction}
   In practice, information systems%
   \footnote{In the sequel, the word ``system'' refers to the phrase ``information system'', to simplify the language a little.}
   may live for many years.
   After they are built, they need to be updated regularly to keep up with changing requirements in a dynamically evolving environment.
   Roughly half of the DevOps~\cite{BassWeberZhu15} teams that responded in a worldwide survey in 2023~\cite{HumanitecDevOps2023} are deploying software more frequently than once per day.
   Obviously, these deployments are mostly updates of existing systems.
   Schema changes cannot always be avoided when updating software, so a {\em schema changing data migration} (SCDM) will be neccessary from time to time.
   For example, adding or removing a column to a table in a relational database adds to the complexity of migrating data.
   Even worse, if a system invariant changes, some of the existing data in the system may violate the new invariant.
   The risk and effort of such data migrations explains why development teams try to avoid schema changes at all costs.

@@

\begin{definition}[Event]
   Let $R \subseteq \Rels$ and $I \subseteq \Rels$ be sets of relations,
   and let $\infsys=\pair{\schema}{\dataset}$ and  $\infsys'=\pair{\schema}{\dataset'}$ be information systems,
t  hen $\infsys\xrightarrow[I]{R} \infsys'$ is an event if and only if:
\begin{align}
   \triple{a}{r}{b}\in\triples_{\dataset}-\triples_{\dataset'}&\Rightarrow\ r\in R\label{eqn:eventUnchanged}\\
   \triple{a}{r}{b}\in\triples_{\dataset'}-\triples_{\dataset}&\Rightarrow\ r\in(R \cup I)\label{eqn:eventInsert}
\end{align}
% was:
% \begin{align}
%    \triples_{\dataset} - (\Triple{\Atoms}{(R \cup I)}{\Atoms}) &= \triples_{\dataset'} - (\Triple{\Atoms}{(R \cup I)}{\Atoms})
% \label{eqn:eventUnchanged}\\
%    \triples_{\dataset} - (\Triple{\Atoms}{R}{\Atoms}) &\subseteq \triples_{\dataset'} - (\Triple{\Atoms}{R}{\Atoms})
% \label{eqn:eventInsert}
% \end{align}
   We say that event $\infsys\xrightarrow[I]{R} \infsys'$ brings the system $\infsys$ to $\infsys'$ by inserting pairs in relations from $I$ and deleting pairs in relations from $R$.
\end{definition}

>>>>

\title{Data Migration under a Changing Schema in Ampersand}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Sebastiaan Joosten\inst{1}\orcidID{0000-0002-6590-6220}\\ \and
Stef Joosten\inst{2,3}\orcidID{0000-0001-8308-0189}}
%
\authorrunning{S. Joosten}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{University of Minnesota, Minneapolis, USA \and Open Universiteit, Heerlen, the Netherlands\\ 
\email{stef.joosten@ou.nl} \and
Ordina NV, Nieuwegein, the Netherlands}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
   Software generators compile a specification into a working information system.
   This can help increase the frequency of releases and their reliability
   because the generator saves on development time and prevents human-induced mistakes.
   However, many generators do not support data migrations.
   Data migration is necessary when an incremental deployment changes the system's schema.
   As a result, developers tend to avoid migrations or migrate data ``by hand''.

@@

   This paper defines a migration process that preserves the semantics of data by means of invariants (constraints)
   with zero downtime for users.
   The migration process is based on assumptions and requirements that aim to capture a large class of data migrations in practice.
   This paper focuses on correctness of the migration. Efficiency of the migration is outside its scope.
\keywords{Generative software \and Incremental software deployment \and Data migration \and Relation algebra \and Ampersand \and Schema change \and Invariants \and Zero downtime}
\end{abstract}

\section{Introduction}
\label{sct:Introduction}
   In practice, information systems%
   \footnote{In the sequel, the word ``system'' refers to the phrase ``information system'', to simplify the language a little.}
   may live for many years.
   After they are built, they need to be updated regularly to keep up with changing requirements in a dynamically evolving environment.
   Roughly half of the DevOps~\cite{BassWeberZhu15} teams that responded in a worldwide survey in 2023~\cite{HumanitecDevOps2023} are deploying software more frequently than once per day.
   Obviously, these deployments are mostly updates of existing systems.
   Schema changes cannot always be avoided when updating software, so a {\em schema changing data migration} (SCDM) will be neccessary from time to time.
   For example, adding or removing a column to a table in a relational database adds to the complexity of migrating data.
   Even worse, if a system invariant changes, some of the existing data in the system may violate the new invariant.
   The risk and effort of such data migrations explains why development teams try to avoid schema changes.

@@

\begin{definition}[Event]
   Let $R \subseteq \Rels$ and $I \subseteq \Rels$ be sets of relations,
   and let $\infsys=\pair{\schema}{\dataset}$ and  $\infsys'=\pair{\schema}{\dataset'}$ be information systems,
   then $\infsys\xrightarrow[I]{R} \infsys'$ is an event if and only if:
\begin{align}
   \triple{a}{r}{b}\in\triples_{\dataset}-\triples_{\dataset'}&\Rightarrow\ r\in R\label{eqn:eventUnchanged}\\
   \triple{a}{r}{b}\in\triples_{\dataset'}-\triples_{\dataset}&\Rightarrow\ r\in(R \cup I)\label{eqn:eventInsert}
\end{align}
% was:
% \begin{align}
%    \triples_{\dataset} - (\Triple{\Atoms}{(R \cup I)}{\Atoms}) &= \triples_{\dataset'} - (\Triple{\Atoms}{(R \cup I)}{\Atoms})
% \label{eqn:eventUnchanged}\\
%    \triples_{\dataset} - (\Triple{\Atoms}{R}{\Atoms}) &\subseteq \triples_{\dataset'} - (\Triple{\Atoms}{R}{\Atoms})
% \label{eqn:eventInsert}
% \end{align}
   We say that event $\infsys\xrightarrow[I]{R} \infsys'$ brings the system $\infsys$ to $\infsys'$ by inserting pairs in relations from $I$ and deleting pairs in relations from $R$.
\end{definition}
